Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-17 analysis/paper/paper.qmd

Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). @lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to data availability, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads. In general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).

The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, we found that variability decreased during the LGM in comparison to the previous period. This could be explained by the overall massive decline in stemmed points usage. We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites). However, as predicted, TCSA values in inland areas such as CT of vegetation zones show higher variability. Based on the results, we assume that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. Our results support this claim by showing the dense distribution of stemmed point in certain environment.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did. The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites panel A & D). Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃. Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.

Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones. Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).

>>>>

Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips. None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). @lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to the unavailability of their data, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads. In general, we found a wider range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.

TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explored the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). We speculate this might be due to the higher availability of raw materials such as shale on hornfels on the landscape. Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).

Our results shows that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period. This could be explained by the overall decline in stemmed points usage.

We found that stemmed points are most frequently located in a small number of eco-regional zones such as the Central Temperature zone with few clears pattern in TCSA function across the zones (@fig-gis-tcsa-sites). As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability. Based on these results, we assume that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were likely more adapted for particular environments, rather than specific climate contexts. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. We similarly found that stemmed points are more densly distributed in certain environments in South Korea, perhaps also relating to the distribution of fauna.

Overall, our temporal and spatial patterns of TCSA show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones. We confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
